---
title: "Stage 02 — Data Profiling"
author: "Hemanth Sai Charagundla"
output: html_document
---------------------

## 1. Purpose

This stage performs comprehensive profiling of the raw CPC18 dataset.

The profiling describes the dataset structure, variables, missing values, duplicates, numeric and categorical variables, and CPC18-specific characteristics without modifying the raw data.

## 2. Execution Order

The profiling stage is executed after data loading.

```text
01_data_loading.R
02_data_profiling.R
```

The raw CPC18 dataset is loaded first and then passed to the profiling functions.

## 3. Load Required Packages

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "..")

library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
```

## 4. Load Data and Run Profiling

```{r profiling, message=FALSE, warning=FALSE}
source("R/01_data_loading.R")
source("R/02_data_profiling.R")

cpc18_profile <- profile_cpc18(cpc18_raw)
```

## 5. Dataset Overview

The dataset overview summarizes the number of rows, columns, total cells, missing values, complete rows, and duplicate rows.

```{r cpc18-overview}
cpc18_profile$overview
```

## 6. Variable Profile

The variable profile records the R class, missing values, missing-value percentage, and number of unique observed values for each variable.

```{r cpc18-variables}
cpc18_profile$variables
```

## 7. Missing Values

Variables containing missing values are identified without applying any imputation or removal.

```{r cpc18-missing}
cpc18_profile$variables %>%
  filter(missing > 0)
```

The missing-value pattern is retained as observed at this stage. No imputation or removal is performed during profiling.

## 8. Duplicate Rows

The duplicate-row count is included in the dataset overview.

```{r duplicates}
tibble(
  dataset = "CPC18",
  duplicate_rows = cpc18_profile$overview$duplicate_rows
)
```

## 9. Numeric Variable Summary

The numeric profile provides the mean, standard deviation, minimum, median, and maximum for numeric variables.

```{r cpc18-numeric}
cpc18_profile$numeric_summary
```

## 10. Categorical Variable Summary

The categorical profile shows the observed levels and their frequencies for character and factor variables.

```{r cpc18-categorical}
cpc18_profile$categorical_summary
```

## 11. CPC18-Specific Profile

### Subjects, Games, Sets, Trials and Orders

```{r cpc18-structure}
list(
  subjects = cpc18_profile$cpc18$subjects,
  games = cpc18_profile$cpc18$games,
  sets = cpc18_profile$cpc18$sets,
  trials = cpc18_profile$cpc18$trials,
  orders = cpc18_profile$cpc18$orders
)
```

### Choice Distribution

```{r cpc18-choice}
cpc18_profile$cpc18$choices
```

### Feedback Distribution

```{r cpc18-feedback}
cpc18_profile$cpc18$feedback
```

### Ambiguity Distribution

```{r cpc18-ambiguity}
cpc18_profile$cpc18$ambiguity
```

### Correlation Distribution

```{r cpc18-correlation}
cpc18_profile$cpc18$correlation
```

### Location Distribution

```{r cpc18-location}
cpc18_profile$cpc18$location
```

### Gender Distribution

```{r cpc18-gender}
cpc18_profile$cpc18$gender
```

### Condition Distribution

```{r cpc18-condition}
cpc18_profile$cpc18$condition
```

### Button Distribution

```{r cpc18-button}
cpc18_profile$cpc18$button
```

### Lottery Shape — Option A

```{r cpc18-lottery-a}
cpc18_profile$cpc18$lottery_shape_a
```

### Lottery Shape — Option B

```{r cpc18-lottery-b}
cpc18_profile$cpc18$lottery_shape_b
```

### Reaction-Time Availability

```{r cpc18-rt}
cpc18_profile$cpc18$reaction_time
```

Reaction time is profiled separately because its availability is structured within CPC18. It is not treated as an ordinary missing-value field at this stage.

### Reaction-Time Availability by Set

```{r cpc18-rt-set}
cpc18_profile$cpc18$reaction_time_by_set
```

### Subject-Level Observations

```{r cpc18-subject-observations}
cpc18_profile$cpc18$subject_observations
```

### Game-Level Observations

```{r cpc18-game-observations}
cpc18_profile$cpc18$game_observations
```

## 12. Profiling Summary

The profiling stage establishes the structure and characteristics of the raw CPC18 dataset before any cleaning, validation, transformation, or feature engineering is performed.

The results from this stage will guide the subsequent data-quality and validation stage.