---
title: "Stage 02 — Data Profiling"
author: "Hemanth Sai Charagundla"
output: html_document
---

## 1. Purpose

This stage performs comprehensive profiling of the raw CPC18 and choices13k datasets.

The profiling describes the structure, variables, missing values, duplicates, numeric variables, categorical variables, and dataset-specific characteristics without modifying the raw data.

## 2. Load Required Packages

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "..")

library(dplyr)
library(purrr)
library(tibble)
```

## 3. Load Data and Run Profiling

```{r profiling, message=FALSE, warning=FALSE}
source("R/01_data_loading.R")
source("R/02_data_profiling.R")

cpc18_profile <- profile_cpc18(cpc18_raw)
choices13k_profile <- profile_choices13k(choices13k_raw)
```

## 4. Dataset Overview

### CPC18

```{r cpc18-overview}
cpc18_profile$overview
```

### choices13k

```{r choices13k-overview}
choices13k_profile$overview
```

## 5. Variable Profile

The variable profile records the data class, missing values, missing-value percentage, and number of unique observed values for each variable.

### CPC18

```{r cpc18-variables}
cpc18_profile$variables
```

### choices13k

```{r choices13k-variables}
choices13k_profile$variables
```

## 6. Missing Values

### CPC18

```{r cpc18-missing}
cpc18_profile$variables %>%
  filter(missing > 0)
```

### choices13k

```{r choices13k-missing}
choices13k_profile$variables %>%
  filter(missing > 0)
```

The missing-value pattern is retained as observed at this stage. No imputation or removal is performed during profiling.

## 7. Duplicate Rows

The duplicate-row count is included in the dataset overview.

```{r duplicates}
tibble(
  dataset = c("CPC18", "choices13k"),
  duplicate_rows = c(
    cpc18_profile$overview$duplicate_rows,
    choices13k_profile$overview$duplicate_rows
  )
)
```

## 8. Numeric Variable Summary

### CPC18

```{r cpc18-numeric}
cpc18_profile$numeric_summary
```

### choices13k

```{r choices13k-numeric}
choices13k_profile$numeric_summary
```

## 9. Categorical Variable Summary

### CPC18

```{r cpc18-categorical}
cpc18_profile$categorical_summary
```

### choices13k

```{r choices13k-categorical}
choices13k_profile$categorical_summary
```

## 10. CPC18-Specific Profile

### Subjects, Games, Sets, Trials and Orders

```{r cpc18-structure}
list(
  subjects = cpc18_profile$cpc18$subjects,
  games = cpc18_profile$cpc18$games,
  sets = cpc18_profile$cpc18$sets,
  trials = cpc18_profile$cpc18$trials,
  orders = cpc18_profile$cpc18$orders
)
```

### Choice Distribution

```{r cpc18-choice}
cpc18_profile$cpc18$choices
```

### Feedback Distribution

```{r cpc18-feedback}
cpc18_profile$cpc18$feedback
```

### Ambiguity Distribution

```{r cpc18-ambiguity}
cpc18_profile$cpc18$ambiguity
```

### Correlation Distribution

```{r cpc18-correlation}
cpc18_profile$cpc18$correlation
```

### Reaction-Time Availability

```{r cpc18-rt}
cpc18_profile$cpc18$reaction_time
```

Reaction time is profiled separately because its availability is structured within CPC18. It is not treated as an ordinary missing-value field at this stage.

### Reaction-Time Availability by Set

```{r cpc18-rt-set}
cpc18_profile$cpc18$reaction_time_by_set
```

### Subject-Level Observations

```{r cpc18-subject-observations}
cpc18_profile$cpc18$subject_observations
```

### Game-Level Observations

```{r cpc18-game-observations}
cpc18_profile$cpc18$game_observations
```

## 11. choices13k-Specific Profile

### Number of Problems

```{r choices13k-problems}
choices13k_profile$choices13k$problems
```

### Feedback Distribution

```{r choices13k-feedback}
choices13k_profile$choices13k$feedback
```

### Block Distribution

```{r choices13k-blocks}
choices13k_profile$choices13k$blocks
```

### Ambiguity Distribution

```{r choices13k-ambiguity}
choices13k_profile$choices13k$ambiguity
```

### Correlation Distribution

```{r choices13k-correlation}
choices13k_profile$choices13k$correlation
```

### Problem-Level Observations

```{r choices13k-problem-observations}
choices13k_profile$choices13k$problem_observations
```

### bRate

```{r choices13k-brate}
choices13k_profile$choices13k$b_rate
```

### Standardized bRate

```{r choices13k-brate-std}
choices13k_profile$choices13k$b_rate_std
```

## 12. Profiling Summary

The profiling stage establishes the structure and characteristics of both raw datasets before any cleaning or transformation is performed.

#### The results from this stage will be used to guide the data-cleaning, validation, and analytical dataset construction stages.