---
title: "Stage 03 — Data Cleaning and Validation"
author: "Hemanth Sai Charagundla"
output: html_document
---

# 1. Objective

The objective of this stage is to construct validated analytical versions of the CPC18 and choices13k datasets while preserving the original raw data.

# 2. Prerequisites and Execution Order

Stage 03 depends on the raw datasets loaded by `R/01_data_loading.R`.

The required execution order is:

1. `R/01_data_loading.R` — loads and standardises the raw datasets in memory.
2. `R/03_data_cleaning.R` — validates, cleans, and saves the processed datasets and validation reports.
3. `reports/03_data_cleaning.Rmd` — documents the cleaning and validation results.

`R/02_data_profiling.R` is a separate profiling stage and does not need to be executed as a prerequisite for the cleaning script itself.

The Stage 03 report is located inside the `reports/` directory, so the project root is explicitly set as the working directory before the pipeline scripts are sourced.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "..")

library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
```

The following pipeline scripts are required for this report:

```{r run-cleaning, message=FALSE, warning=FALSE}
source("R/01_data_loading.R")
source("R/03_data_cleaning.R")
```

`R/01_data_loading.R` must run first because `R/03_data_cleaning.R` expects the objects `cpc18_raw` and `choices13k_raw` to already exist in memory.

The cleaning script then creates:

- `cpc18_clean`
- `choices13k_clean`
- `cpc18_cleaning_report`
- `choices13k_cleaning_report`
- `cleaning_summary`

and writes the processed datasets and validation reports to `data/processed/`.

# 3. Cleaning Principles

The cleaning stage follows several principles.

1. **Raw data preservation**  
   Original source files remain unchanged.

2. **No unnecessary row removal**  
   Observations are not removed merely because values are unusual.

3. **Missingness is preserved unless a justified transformation is required**  
   In particular, missing reaction-time values in CPC18 are retained because reaction time was not collected for all experimental sets.

4. **Schema validation precedes transformation**  
   Expected variables are checked before the cleaning process is applied.

5. **Logical validation is separated from analytical decisions**  
   A value that is unusual is not automatically treated as erroneous.

6. **Row-count integrity is explicitly checked**  
   Cleaning should not silently remove observations.

# 4. Dataset Schema Validation

The cleaning script defines expected schemas for both datasets.

For CPC18, the expected variables include participant identifiers, experimental conditions, lottery characteristics, choices, payoffs, reaction time, feedback, and trial structure.

For choices13k, the expected variables include problem characteristics, feedback, block information, lottery characteristics, ambiguity, correlation, and aggregate choice-rate measures.

The function `validate_columns()` checks whether all required variables are present before cleaning proceeds.

# 5. CPC18 Cleaning

## 5.1 Type Standardisation

The CPC18 cleaning function standardises variables according to their semantic roles.

Identifiers and experimental indices are stored as integers, probabilities and numerical outcome variables as numeric, binary experimental variables as integer indicators, and reaction time as numeric.

Reaction-time availability is calculated from the original reaction-time field and recorded in the cleaning validation report.

```{r cpc18-type-validation}
cpc18_cleaning_report$rt_by_set
```

## 5.2 Missing Values

The CPC18 profiling stage identified missing values only in the reaction-time variable.

```{r cpc18-missing}
tibble(
  total_observations = nrow(cpc18_clean),
  rt_available = sum(!is.na(cpc18_clean$rt)),
  rt_missing = sum(is.na(cpc18_clean$rt)),
  rt_available_pct = mean(!is.na(cpc18_clean$rt)) * 100,
  rt_missing_pct = mean(is.na(cpc18_clean$rt)) * 100
)
```

The observed reaction-time pattern is structured by experimental set rather than randomly distributed across the dataset.

Therefore, reaction-time missingness is preserved rather than imputed during this stage.

## 5.3 Duplicate Validation

```{r cpc18-duplicates}
cpc18_cleaning_report$duplicates
```

The raw CPC18 dataset contains no duplicate rows.

The cleaning process therefore does not remove observations on the basis of duplication.

## 5.4 Logical Validation

The cleaning process checks:

- participant age;
- probabilities;
- binary choice values;
- ambiguity indicators;
- feedback indicators;
- correlation values;
- trial numbers;
- order positions;
- block values;
- negative reaction times.

```{r cpc18-validation}
cpc18_validation_checks <- tibble(
  check = c(
    "Invalid age",
    "Invalid probabilities",
    "Invalid choice",
    "Invalid ambiguity",
    "Invalid feedback",
    "Invalid correlation",
    "Invalid trial",
    "Invalid order",
    "Invalid block",
    "Negative reaction time"
  ),
  invalid_values = c(
    cpc18_cleaning_report$invalid_age,
    cpc18_cleaning_report$invalid_probabilities,
    cpc18_cleaning_report$invalid_choice,
    cpc18_cleaning_report$invalid_ambiguity,
    cpc18_cleaning_report$invalid_feedback,
    cpc18_cleaning_report$invalid_correlation,
    cpc18_cleaning_report$invalid_trial,
    cpc18_cleaning_report$invalid_order,
    cpc18_cleaning_report$invalid_block,
    cpc18_cleaning_report$negative_rt
  )
)

cpc18_validation_checks
```

These checks allow structurally invalid values to be identified before downstream analysis.

# 6. CPC18 Experimental Structure Validation

The dataset contains repeated observations organised around subjects, games, and trials.

## 6.1 Subject Structure

```{r cpc18-subject}
cpc18_cleaning_report$subject_structure
```

The cleaning validation confirms the observation structure at the participant level.

## 6.2 Game Structure

```{r cpc18-game}
cpc18_cleaning_report$game_structure
```

Game-level observation counts are checked to identify unexpected structural inconsistencies.

## 6.3 Reaction Time by Experimental Set

```{r cpc18-rt-structure}
cpc18_cleaning_report$rt_by_set
```

This check is particularly important because reaction time is not available for every experimental set.

The structured missingness will therefore be treated explicitly in later reaction-time analysis rather than being treated as ordinary missing data.

# 7. choices13k Cleaning

## 7.1 Type Standardisation

The choices13k cleaning function standardises identifiers, experimental indices, probabilities, lottery characteristics, binary variables, and choice-rate variables according to their intended data types.

## 7.2 Missing Values

The profiling stage showed that choices13k contains no missing values.

```{r choices13k-missing}
choices13k_cleaning_report$missing_values
```

No imputation is therefore required during this cleaning stage.

## 7.3 Duplicate Validation

```{r choices13k-duplicates}
choices13k_cleaning_report$duplicates
```

The dataset contains no duplicate rows.

## 7.4 Logical Validation

The cleaning process checks:

- probability bounds;
- choice-rate bounds;
- standardised choice-rate values;
- correlation values;
- block values;
- valid values of `n`.

```{r choices13k-validation}
choices13k_validation_checks <- tibble(
  check = c(
    "Invalid probabilities",
    "Invalid choice rates",
    "Invalid standardised choice rates",
    "Invalid correlation",
    "Invalid block",
    "Invalid n"
  ),
  invalid_values = c(
    choices13k_cleaning_report$invalid_probabilities,
    choices13k_cleaning_report$invalid_choice_rates,
    choices13k_cleaning_report$invalid_standardised_choice_rates,
    choices13k_cleaning_report$invalid_correlation,
    choices13k_cleaning_report$invalid_block,
    choices13k_cleaning_report$invalid_n
  )
)

choices13k_validation_checks
```

# 8. Problem-Level Structure

The choices13k dataset contains repeated observations for some problems.

```{r choices13k-problems}
choices13k_cleaning_report$problem_structure
```

This structure is retained because repeated observations may be analytically meaningful.

No observations are removed solely because a problem appears more than once.

# 9. Row-Count Integrity

A critical requirement of the cleaning process is that the number of observations should not change unexpectedly.

```{r cleaning-summary}
cleaning_summary
```

The final datasets are compared with their corresponding raw datasets using:

- original row count;
- cleaned row count;
- number of rows removed;
- remaining missing values;
- remaining duplicate rows.

```{r cleaning-integrity}
cleaning_integrity <- tibble(
  dataset = c("CPC18", "choices13k"),
  original_rows = c(
    nrow(cpc18_raw),
    nrow(choices13k_raw)
  ),
  cleaned_rows = c(
    nrow(cpc18_clean),
    nrow(choices13k_clean)
  ),
  rows_removed = c(
    nrow(cpc18_raw) - nrow(cpc18_clean),
    nrow(choices13k_raw) - nrow(choices13k_clean)
  ),
  duplicates_after_cleaning = c(
    sum(duplicated(cpc18_clean)),
    sum(duplicated(choices13k_clean))
  ),
  missing_values_after_cleaning = c(
    sum(is.na(cpc18_clean)),
    sum(is.na(choices13k_clean))
  )
)

cleaning_integrity
```

# 10. Cleaning Outcome

The cleaning stage produces two processed datasets:

- `CPC18_clean`
- `choices13k_clean`

The original raw datasets remain untouched.

The cleaning process is intentionally conservative. It standardises data types and validates structural and logical constraints without prematurely deleting observations or altering behavioural variables.

In particular, CPC18 reaction-time missingness is retained for dedicated reaction-time analysis.

# 11. Analytical Implications

Several observations from this stage are important for subsequent analysis.

### CPC18

CPC18 contains repeated observations within participants and games. Therefore, later statistical and machine-learning analyses must account for the repeated-measures structure where appropriate.

Reaction time is available only for a subset of the experimental data. Consequently, reaction-time analyses will operate on the valid RT subset rather than using arbitrary imputation.

### choices13k

choices13k is a population-level/problem-level dataset with aggregate choice-rate variables. It should therefore be analysed according to its own structure rather than treated as another individual-level version of CPC18.

The two datasets will remain analytically separate until the later cross-dataset comparison stage.

# 12. Output Files

The cleaning stage writes the processed datasets to:

```text
data/processed/CPC18_clean.csv
data/processed/choices13k_clean.csv
```

Validation reports are also stored as RDS objects for reproducibility:

```text
data/processed/CPC18_cleaning_report.rds
data/processed/choices13k_cleaning_report.rds
data/processed/cleaning_summary.rds
```

# 13. Conclusion

The cleaning stage establishes validated analytical versions of both datasets while preserving the original source data and important experimental structure.

The resulting datasets are ready for the next stage: **data processing and analytical dataset construction**.

#### Further transformations, derived variables, and decision-focused features are intentionally deferred to subsequent stages.