---
title: "Stage 03 — Data Cleaning"
author: "Hemanth Sai Charagundla"
output: html_document
---------------------

## 1. Purpose

This stage creates a clean analytical copy of the raw CPC18 dataset while preserving the original source file.

The cleaning process assigns meaningful variable names, standardizes character values, validates expected data types, records reaction-time availability, and removes exact duplicate rows if present.

## 2. Load Required Packages

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "..")

library(dplyr)
library(data.table)
```

## 3. Load and Clean Dataset

```{r data-cleaning, message=FALSE, warning=FALSE}
source("R/01_data_loading.R")
source("R/03_data_cleaning.R")
```

The cleaning script creates the in-memory analytical dataset `cpc18_cleaned` and saves it to:

```text
data/processed/cpc18_cleaned.csv
```

## 4. Meaningful Variable Names

The original CPC18 variable names are replaced with descriptive names based on the CPC18 data dictionary.

Examples include:

* `ha` → `expected_value_a`
* `p_ha` → `probability_a`
* `hb` → `expected_value_b`
* `p_hb` → `probability_b`
* `amb` → `ambiguous_b`
* `corr` → `payoff_correlation`
* `b` → `chose_b`
* `rt` → `reaction_time_ms`
* `feedback` → `full_feedback`
* `block` → `time_block`

This makes the analytical dataset easier to interpret while leaving the original raw file unchanged.

## 5. Data Type Validation

The cleaning process defines the expected data type for each CPC18 variable and compares it with the observed type.

If an unexpected type is detected, the script stops and reports the affected variable.

This prevents incorrect data types from being silently carried into subsequent analysis.

## 6. Reaction-Time Availability

A Boolean `rt_available` variable is created from the reaction-time field.

```{r rt-availability}
tibble(
  rt_available = c(TRUE, FALSE),
  observations = c(
    sum(cpc18_cleaned$rt_available),
    sum(!cpc18_cleaned$rt_available)
  )
)
```

The resulting indicator distinguishes observations where reaction time was recorded from those where it was unavailable.

Reaction-time values themselves are not imputed or removed during this stage.

## 7. Duplicate Removal

```{r duplicates}
tibble(
  original_rows = nrow(cpc18_raw),
  cleaned_rows = nrow(cpc18_cleaned),
  rows_removed = nrow(cpc18_raw) - nrow(cpc18_cleaned)
)
```

The exact duplicate check is performed using `distinct()`.

The profiling stage identified no duplicate rows in the raw CPC18 dataset, so no observations are expected to be removed by this operation.

## 8. Cleaning Confirmation

The cleaned dataset contains:

```{r confirmation}
tibble(
  rows = nrow(cpc18_cleaned),
  columns = ncol(cpc18_cleaned),
  rt_available = sum(cpc18_cleaned$rt_available),
  rt_unavailable = sum(!cpc18_cleaned$rt_available)
)
```

The cleaned CPC18 dataset retains the original observations while adding the standardized analytical variable names and the reaction-time availability indicator.

The raw source file remains unchanged.

## 9. Output

The cleaned dataset is saved as:

```text
data/processed/cpc18_cleaned.csv
```

Further logical validation, analytical transformations, and feature engineering are performed in subsequent stages.