---
title: "Stage 04 — Data Validation"
author: "Hemanth Sai Charagundla"
output: html_document
---------------------

## 1. Purpose

This stage validates the logical consistency of the cleaned CPC18 analytical dataset.

The validation checks categorical values, binary coding, numeric ranges, trial and time-block consistency, payoff relationships, and reaction-time validity.

Validation does not modify or remove observations. Any detected anomalies are reported through the validation results.

## 2. Prerequisites and Execution Order

Stage 04 depends on the cleaned CPC18 dataset produced by `R/03_data_cleaning.R`.

The required execution order is:

1. `R/01_data_loading.R` — loads the raw CPC18 dataset.
2. `R/03_data_cleaning.R` — creates and saves the cleaned CPC18 dataset.
3. `R/04_data_validation.R` — validates the cleaned CPC18 dataset and generates validation results.

## 3. Load Required Packages

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "..")

library(dplyr)
```

## 4. Load and Validate Dataset

```{r validation, message=FALSE, warning=FALSE}
source("R/04_data_validation.R")
```

The validation script loads:

`data/processed/cpc18_cleaned.csv`

and produces the validation results table:

`validation_results`.

## 5. Validation Checks

### 5.1 Categorical Values

The following categorical variables are checked against their expected CPC18 values:

* `location`
* `gender`
* `condition`
* `lottery_shape_a`
* `lottery_shape_b`
* `selected_button`

### 5.2 Binary Variables

The following variables are validated to contain only their expected binary coding:

* `ambiguous_b`
* `chose_b`
* `full_feedback`

### 5.3 Numeric Ranges

The following numerical constraints are validated:

* `probability_a` and `probability_b` must be between 0 and 1.
* `payoff_correlation` must be `-1`, `0`, or `1`.
* `game_order` must be between 1 and 30.
* `trial_number` must be between 1 and 25.
* `time_block` must be between 1 and 5.

### 5.4 Trial and Block Consistency

The time-block assignment is checked against the trial number using the five-trial block structure.

### 5.5 Payoff Consistency

The validation checks whether:

* `realized_payoff` corresponds to the selected option.
* `forgone_payoff` corresponds to the option not selected.

### 5.6 Reaction-Time Validation

Recorded reaction times are checked for negative values.

The `rt_available` indicator is also checked against the presence or absence of `reaction_time_ms`.

Missing reaction times are not treated as invalid observations.

## 6. Validation Results

```{r validation-results}
validation_results
```

The validation results contain:

* `check` — validation rule that was evaluated.
* `invalid_rows` — number of observations that failed the check.
* `status` — `PASS` when no invalid observations were detected and `FLAG` when observations require attention.

## 7. Validation Summary

```{r validation-summary}
validation_results %>%
  count(status)
```

The complete validation output is saved to:

`data/processed/cpc18_validation_results.csv`

## 8. Interpretation

The validation stage provides a logical quality check before downstream analysis and feature construction.

A `PASS` indicates that no observations violated the corresponding validation rule.

The validation process does not automatically remove flagged observations. Any anomaly identified at this stage must be investigated and its treatment documented before subsequent analysis.

## 9. Output

The cleaned analytical dataset remains:

`data/processed/cpc18_cleaned.csv`

Validation results are stored separately as:

`data/processed/cpc18_validation_results.csv`

The raw CPC18 source data remains unchanged.

## 10. Next Stage

After data validation, the project proceeds to the next analytical stage using the validated CPC18 dataset.