---
title: "Stage 05 — Feature Engineering"
author: "Hemanth Sai Charagundla"
output: html_document
---------------------

## 1. Purpose

This stage creates mathematically and analytically meaningful features from the validated CPC18 dataset.

The engineered features represent expected payoff, probability differences, payoff structure, lottery structure, and experimental conditions relevant to human decision-making under uncertainty.

Features are constructed without modifying the original cleaned variables.

## 2. Prerequisites and Execution Order

Stage 05 depends on the validated CPC18 dataset produced by `R/04_data_validation.R`.

The required execution order is:

1. `R/01_data_loading.R` — loads the raw CPC18 dataset.
2. `R/03_data_cleaning.R` — creates the cleaned CPC18 dataset.
3. `R/04_data_validation.R` — validates the cleaned CPC18 dataset.
4. `R/05_feature_engineering.R` — creates mathematically meaningful analytical features.
5. `reports/05_feature_engineering.Rmd` — documents the feature-engineering process and results.

The project root is explicitly set as the working directory before the pipeline scripts are sourced.

## 3. Load Required Packages

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "..")

library(dplyr)
library(data.table)
library(tibble)
```

## 4. Load and Create Features

```{r feature-engineering, message=FALSE, warning=FALSE}
source("R/05_feature_engineering.R", local = TRUE)
```

The feature-engineering script loads:

`data/processed/cpc18_cleaned.csv`

and creates the feature-enhanced dataset:

`cpc18_features`.

The resulting dataset contains the observations and variables generated by the feature-engineering script.

## 5. Expected Payoff Features

The complete expected payoff of each option is calculated using the probability of receiving the high-lottery payoff and the low payoff:

$$
E(A) = p_A H_A + (1-p_A)L_A
$$

$$
E(B) = p_B H_B + (1-p_B)L_B
$$

This produces:

* `expected_payoff_a`
* `expected_payoff_b`

The expected-payoff difference is then calculated as:

$$
\Delta E = E(B) - E(A)
$$

producing `expected_payoff_diff`.

A corresponding absolute difference,

`absolute_expected_payoff_diff`,

captures the magnitude of the expected-payoff gap without its direction.

## 6. Probability and Payoff Features

The probability difference is defined as:

$$
\Delta p = p_B - p_A
$$

producing `probability_diff`.

Its absolute value is stored as:

`absolute_probability_diff`.

The difference in low payoffs is calculated as:

`low_payoff_diff`.

The difference between the expected values of the high lotteries is calculated as:

`high_lottery_ev_diff`.

## 7. Payoff Spread Features

For each option, the difference between the high-lottery expected value and its low payoff is calculated:

$$
Spread_A = H_A - L_A
$$

$$
Spread_B = H_B - L_B
$$

These produce:

* `payoff_spread_a`
* `payoff_spread_b`

Their difference is stored as:

`payoff_spread_diff`.

These variables describe payoff structure and are not labelled as formal measures of risk.

## 8. Lottery Structure

The difference in the number of possible lottery outcomes is calculated as:

`lottery_outcomes_diff`.

This represents the relative difference in lottery outcome counts between options A and B.

The number of outcomes is retained as a structural characteristic rather than being interpreted as a direct measure of risk.

## 9. Experimental Features

The following existing CPC18 variables are retained as relevant pre-decision analytical features:

* `ambiguous_b`
* `payoff_correlation`
* `full_feedback`
* `game_order`
* `trial_number`
* `time_block`

These variables represent experimental conditions or temporal position and therefore have direct analytical relevance to the decision-making questions.

## 10. Pre-Decision Feature Set

The project defines a set of 18 variables for the pre-decision analytical feature set.

```{r feature-list}
pre_decision_features
```

These variables are intended for subsequent decision-focused analysis and leakage-safe modelling.

## 11. Leakage Consideration

Post-decision outcome variables are retained in the feature-enhanced dataset because they may be useful for later behavioural and outcome analysis.

However, the following variables are not included in the pre-decision feature set:

* `realized_payoff`
* `forgone_payoff`
* `payoff_a`
* `payoff_b`

These variables contain information generated from the decision outcome and therefore require separate treatment when constructing pre-decision choice-prediction models.

## 12. Feature-Engineering Confirmation

```{r feature-confirmation}
tibble(
  rows = nrow(cpc18_features),
  columns = ncol(cpc18_features),
  engineered_features = length(pre_decision_features)
)
```

The feature-enhanced CPC18 dataset contains the generated observations and variables, including the defined pre-decision analytical feature set.

## 13. Output

The feature-enhanced dataset is saved as:

`data/processed/cpc18_features.csv`

The original raw and cleaned datasets remain unchanged.

## 14. Next Stage

The next stage will perform decision-focused exploratory data analysis using the validated CPC18 variables and the newly engineered features.